Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/appendix.qmd
Use right appendix cross referencing

---
title: 'Online appendix for "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"'
short-title: 'Online appendix for "Derogations and Backsliding"'
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: 2024-06-30
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
bibliography: references.json
csl: csl/apa.csl
link-citations: true
crossref:
  custom:
    - kind: float
      key: appfig
      latex-env: appfig
      reference-prefix: Figure A
      space-before-numbering: false
      latex-list-of-description: Appendix Figure
    - kind: float
      key: apptbl
      latex-env: apptbl
      reference-prefix: Table A
      space-before-numbering: false
      latex-list-of-description: Appendix Table
apptbl-cap-location: top
toc: true
toc-depth: 2
---

@@

```{r tbl-data-sources}
#| tbl-cap: Sources for outcomes and explanatory variables
#| include: true

@@

>>>>

---
title: 'Online appendix for "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"'
short-title: 'Online appendix for "Derogations and Backsliding"'
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: 2024-11-25
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
bibliography: references.json
csl: csl/apa.csl
link-citations: true
crossref:
  custom:
    - kind: float
      key: appfig
      latex-env: appfig
      reference-prefix: Figure A
      space-before-numbering: false
      latex-list-of-description: Appendix Figure
    - kind: float
      key: apptbl
      latex-env: apptbl
      reference-prefix: Table A
      space-before-numbering: false
      latex-list-of-description: Appendix Table
apptbl-cap-location: top
toc: true
toc-depth: 2
---

@@

::: {#apptbl-data-sources}

```{r table-data-sources}
#| include: true

@@

Sources for outcomes and explanatory variables

### 2024-11-25 manuscript/manuscript.qmd
Use right appendix cross referencing

We use weekly data from March 11, 2020 to June 30, 2021 for 139 countries during the first 69 weeks of the pandemic. @tbl-data-sources lists the main variables we used across our models. To measure democratic backsliding, we use the Pandemic Backsliding Index (PanBack) from the Varieties of Democracy (V-Dem) project [@vdem-v12; @vdem-pandem]. PanBack measures the risk of democratic backsliding specifically during the pandemic, assessing how state responses to the pandemic violated democratic standards, based on the severity of different kinds of human rights violations. PanBack ranges from 0 to 1, with high values representing greater risk of backsliding. Following V-Dem, we dichotomize values and consider PanBack scores above 0.3 as high risk.

@@

We use a Bayesian approach to explore the uncertainty associated with state behavior during the pandemic---we include detailed specifications of our models, priors, and sampling strategy in the appendix. For binary outcomes, we use logistic regression; for categorical outcomes, we use ordered logistic regression. Both of these families of models provide coefficients on a logged odds scale, which can make interpretation difficult. To aid in the interpretation of results, we calculate conditional predicted probabilities by holding all explanatory variables constant (representing a typical country/week) and varying only derogation status and backsliding risk. We then calculate the contrasts between these predicted probabilities to determine the average differences associated with derogations and backsliding. We present these predicted probabilities graphically where possible, and we include the complete log-odds-scale results in Tables A1--A3 in the appendix. Rather than reporting individual point estimates, we report 95% credible intervals. We also report the posterior probability that the predicted contrasts between derogation status and/or backsliding risk are above or below zero.

@@

We first explore whether backsliding states are more likely to issue derogations from international human rights treaties. @fig-derog-coefs presents the odds ratios from a model predicting ICCPR derogations (complete results are in Appendix Table 1). States that experienced pandemic backsliding were nearly four times more likely to file derogations (`r derog_coefs_l$b_panback$y_nice`; `r derog_coefs_l$b_panback$p_d`). This is distinct from states with strong domestic institutions as measured by the rule of law index, which was positively associated with the probability of derogation, but not statistically significant (`r derog_coefs_l$b_v2x_rule$y_nice`; `r derog_coefs_l$b_v2x_rule$p_d`). Looking at the measures of health crisis, a one standard deviation increase in new and cumulative COVID cases is associated with a `r derog_coefs_l$b_new_cases_z$or_pct` and `r derog_coefs_l$b_cumulative_cases_z$or_pct` decrease in the likelihood of derogation, respectively. In contrast, an increase in new and cumulative COVID deaths is associated with a ≈50% higher and a nearly three times greater likelihood of derogation, respectively. Taken together, we see that more severe measures which captured increased intensity of both (1) democratic violations contributing to backsliding and (2) COVID deaths were the indicators that prompted states to issue pandemic-related derogations. Backsliding states were more likely to issue derogations to the ICCPR, limiting legal obligations to international human rights law during the pandemic.

>>>>

We use weekly data from March 11, 2020 to June 30, 2021 for 139 countries during the first 69 weeks of the pandemic. Table A1 in the appendix lists the main variables we used across our models. To measure democratic backsliding, we use the Pandemic Backsliding Index (PanBack) from the Varieties of Democracy (V-Dem) project [@vdem-v12; @vdem-pandem]. PanBack measures the risk of democratic backsliding specifically during the pandemic, assessing how state responses to the pandemic violated democratic standards, based on the severity of different kinds of human rights violations. PanBack ranges from 0 to 1, with high values representing greater risk of backsliding. Following V-Dem, we dichotomize values and consider PanBack scores above 0.3 as high risk.

@@

We use a Bayesian approach to explore the uncertainty associated with state behavior during the pandemic---we include detailed specifications of our models, priors, and sampling strategy in the appendix. For binary outcomes, we use logistic regression; for categorical outcomes, we use ordered logistic regression. Both of these families of models provide coefficients on a logged odds scale, which can make interpretation difficult. To aid in the interpretation of results, we calculate conditional predicted probabilities by holding all explanatory variables constant (representing a typical country/week) and varying only derogation status and backsliding risk. We then calculate the contrasts between these predicted probabilities to determine the average differences associated with derogations and backsliding. We present these predicted probabilities graphically where possible, and we include the complete log-odds-scale results in Tables A2--A4 in the appendix. Rather than reporting individual point estimates, we report 95% credible intervals. We also report the posterior probability that the predicted contrasts between derogation status and/or backsliding risk are above or below zero.

@@

We first explore whether backsliding states are more likely to issue derogations from international human rights treaties. @fig-derog-coefs presents the odds ratios from a model predicting ICCPR derogations (complete results are in Appendix Table 2). States that experienced pandemic backsliding were nearly four times more likely to file derogations (`r derog_coefs_l$b_panback$y_nice`; `r derog_coefs_l$b_panback$p_d`). This is distinct from states with strong domestic institutions as measured by the rule of law index, which was positively associated with the probability of derogation, but not statistically significant (`r derog_coefs_l$b_v2x_rule$y_nice`; `r derog_coefs_l$b_v2x_rule$p_d`). Looking at the measures of health crisis, a one standard deviation increase in new and cumulative COVID cases is associated with a `r derog_coefs_l$b_new_cases_z$or_pct` and `r derog_coefs_l$b_cumulative_cases_z$or_pct` decrease in the likelihood of derogation, respectively. In contrast, an increase in new and cumulative COVID deaths is associated with a ≈50% higher and a nearly three times greater likelihood of derogation, respectively. Taken together, we see that more severe measures which captured increased intensity of both (1) democratic violations contributing to backsliding and (2) COVID deaths were the indicators that prompted states to issue pandemic-related derogations. Backsliding states were more likely to issue derogations to the ICCPR, limiting legal obligations to international human rights law during the pandemic.
